Computational Methods for Principle Component Analysis Applied to Genotype Data

### 2018-03-04 doc/paper.tex
Correct table label.
Add discussion and data tables in paper

\title{Computational Methods for Principle Component Analysis Applied to
  Genotype Data}
\date{}

@@

\subsubsection{Randomized SVD}
In addition to estimating the PC scores of the study individuals,
we have another algorithm that aims at improving the decomposition of the
training data.
For SP, AP, ADT, and OADT, we need to find the covariance matrix of the training
data, which has dimension $n \times n$, and conduct eigendecomosition on it.
However, when $n$ is exceedingly large,
the covariance matrix cannot be loaded into the memory of an average personal
computer.
A randomized SVD (RSVD) algorithm has been developed to handle this issue.
RSVD is slower than standard SVD,
but it does not require loading the whole covariance matrix into the memory.
For the details of the RSVD algorithm, see \cite{halko} and \cite{abraham}.

@@

The predicted PC scores are shown in \Cref{fig:n1000} and \Cref{fig:ukb}.
For the comparison of the runtimes and accuracy for different reference and sample sizes for the simulated data,
see \Cref{fig:nChg} and \Cref{fig:mChg}.
For comparing the methods for the real-life data, see

@@

We find that AP improved the accuracy of SP without increasing the runtimes.
Moreover, OADT improves the speed of ADT without much sacrifice of accuracy.
In addition, randomized SVD is found to be as reliable as standard SVD in
calculating the PC's of the reference group.

@@

\begin{table} 
  \centering
  \begin{tabular}{|l|l|l|}
    \hline
    Method & Reference Complexity & Study Complexity \\ 
    \hline
    Simple Projection (SP) & $\bO[n^2p]$ & $\bO[mkp]$ \\
    \hline
    Randomized Simple Projection (RSP) & $\bO[k^2p]$ & $\bO[mkp]$ \\
    \hline
    Augment-Decompose-Transform (ADT) & $\bO[n^2 p]$ & $\bO[mn(p + n^2)]$ \\
    \hline
    Adjusted Projection (AP) & $\bO[n^2p]$ &  $\bO[mkp]$ \\
    \hline
    Randomized Adjusted Projection (RAP) & $\bO[k^2p]$ &  $\bO[mkp]$ \\
    \hline
    Online Augment-Decompose-Transform (OADT) & $\bO[n^2 p]$ & $\bO[mk(p + k n)]$ \\
    \hline
    Randomized Online Augment-Decompose-Transform (ROADT) & $\bO[k^2 p]$ & $\bO[mk(p + k n)]$ \\
    \hline
  \end{tabular}
  \caption{
    Comparison of computational complexity for different methods.
    Here $p$ is the number of SNPs,
    $n$ is the number of individuals in the reference group,
    $m$ is the number of individuals in the  study group,
    and $k$ is the number of PCs calculated by the corresponding method.
  }
  \label{tbl:cplx}
\end{table}

\begin{table} 
  \centering
  \begin{tabular}{|l|l|l|l|l|}
    \hline
    Method & Reference Runtimes & Study Runtimes & Golden Standard Error \\ 
    \hline
    SP & 411 & 425 & 1.95 \\
    \hline
    RSP & 2504 & 241 & 1.95 \\
    \hline
    ADT & 411 & 29813 & 0.00 \\
    \hline
    AP & 411 & 482 1 & 1.12 \\
    \hline
    RAP & 2505 & 250 & 1.83 \\ 
    \hline
    OADT & 2518 & 103 & 0.40 \\
    \hline
    ROADT & 2518 & 103 & 1.93 \\
    \hline
  \end{tabular}
  \caption{
    Comparison of runtimes and accuracy for different methods applied to the real-life datasets.
    The reference individuals are from the 1000 Genomes dataset with 2492 unrelated individuals.
    The study sample contain 1000 individuals from the UKBioBank dataset.
    There are about 125,000 loci shared by the reference and study data.
    The golden standard error is the square root of the averate squared distance between the study individuals' PC scores predicted by different methods compared to the golden standard result, which in our case is the PC scores predicted by ADT.
  }
  \label{tbl:cplx}
\end{table}

>>>>

\title{Computational Methods for Principle Component Analysis Applied to
  Genotype Data}
\date{}
\author{}
% an example of a edit

@@

\subsubsection{Randomized SVD}

In addition to estimating the PC scores of the study individuals,
we have another algorithm that aims at improving the decomposition of the
training data.
For SP, AP, ADT, and OADT, we need to find the covariance matrix of the training
data, which has dimension $n \times n$, and conduct eigendecomosition on it.
However, when $n$ is exceedingly large,
the covariance matrix cannot be loaded into the memory of an average personal
computer.
A randomized SVD (RSVD) algorithm has been developed to handle this issue.
RSVD is slower than standard SVD,
but it does not require loading the whole covariance matrix into the memory.
For the details of the RSVD algorithm, see \cite{halko} and \cite{abraham}.

Notice that the randomized SVD method on the training data
only calculates the first $k$ eigenvalues.
However, in order to use adjusted projection,
all the eigenvalues of the training data are required.
Thus for the missing eigenvalues,
we predict them by regressing their log-scale values on the ranks
and using the first $k$ eigenvalues for training the linear regression.

@@

The predicted PC scores are shown in \Cref{fig:n1000} and \Cref{fig:ukb}.
For the comparison of the runtimes and accuracy for different reference and sample sizes for the simulated data,
see \Cref{fig:nChg} and \Cref{fig:mChg}.
For comparing the methods for the real-life data, see \Cref{tbl:ukb}.

@@

For the real-life datasets (\Cref{tbl:ukb} and \Cref{fig:ukb}),
we see that ADT, which is the most reliable method,
is the slowest in study runtimes.
By using the online SVD algorithm,
OADT achieved the highest accuracy (excluding ADT, which is the golden standard) and the fastest runtime.
SP is fast, also,
but its accuracy is the lowest.
By using the HDPCA package,
AP reduces the error of AP
and achieved the second highest in accuracy.
The randomized version of SP is as accurate as the standard SP,
which shows that the randomized SVD algorithm gives reliable results,
and that using linear regression to predict the uncalculated reference eigenvalues is satisfactory.
It also reduces the runtime almost by half,
due to not calculating all the eigenvalues.
Similar reduction in study runtime is observed for the randomized version of AP,
but in this case the accuracy is also decreased,
thought it is still higher than that of SP and RSP.
Finally, the randomized OADT is disappointing,
as its error is almost as high as SP and RSP.

The pattern for accuracy is very similar when we look at the simulated data.
(\Cref{tbl:nChg-runtimes-ref}, \Cref{tbl:nChg-runtimes-study},
\Cref{tbl:nChg-accuracy-gold}, \Cref{tbl:nChg-accuracy-ctr},
\Cref{tbl:mChg-runtimes-study}).
A noticeable difference in this case is the accuracy of ROADT.
Here ROADT's accuracy is almost identical to OADT's,
as measured by both the golden standard error and the geographic center err.
However, as the reference size increases,
the accuracies of different methods become closer and closer (\Cref{fig:nChg}).
As for the speed,
AWD's study runtime increases so fast with the reference size
that the study runtimes for the other methods seem to be constant in comparison (\Cref{fig:nChg}).
On the other hand,
the study runtimes appear to increase linearly with respect to the study size (\Cref{fig:mChg}).
This information can be used to predict the runtime for larger study samples.

@@

\begin{table} 
  \centering
  \begin{tabular}{|l|l|l|}
    \hline
    Method & Reference Complexity & Study Complexity \\ 
    \hline
    ADT & $\bO[n^2 p]$ & $\bO[mn(p + n^2)]$ \\
    \hline
    OADT & $\bO[n^2 p]$ & $\bO[mk(p + k n)]$ \\
    \hline
    ROADT & $\bO[k^2 p]$ & $\bO[mk(p + k n)]$ \\
    \hline
    SP & $\bO[n^2p]$ & $\bO[mkp]$ \\
    \hline
    RSP & $\bO[k^2p]$ & $\bO[mkp]$ \\
    \hline
    AP & $\bO[n^2p]$ &  $\bO[mkp]$ \\
    \hline
    RAP & $\bO[k^2p]$ &  $\bO[mkp]$ \\
    \hline
  \end{tabular}
  \caption{
    Comparison of computational complexity for different methods.
    Here $p$ is the number of SNPs,
    $n$ is the number of individuals in the reference group,
    $m$ is the number of individuals in the  study group,
    and $k$ is the number of PCs calculated by the corresponding method.
  }
  \label{tbl:cplx}
\end{table}

\begin{table} 
  \centering
  \begin{tabular}{|l|l|l|l|l|}
    \hline
    Method & Reference Runtimes & Study Runtimes & Golden Standard Error \\ 
    \hline
    ADT & 411 & 29813 & 0.00 \\
    \hline
    OADT & 2518 & 103 & 0.40 \\
    \hline
    ROADT & 2518 & 103 & 1.93 \\
    \hline
    SP & 411 & 425 & 1.95 \\
    \hline
    RSP & 2504 & 241 & 1.95 \\
    \hline
    AP & 411 & 482 & 1.12 \\
    \hline
    RAP & 2505 & 250 & 1.83 \\ 
    \hline
  \end{tabular}
  \caption{
    Comparison of runtimes and accuracy for different methods applied to the real-life datasets.
    The reference individuals are from the 1000 Genomes dataset with 2492 unrelated individuals.
    The study sample contain 1000 individuals from the UKBioBank dataset.
    There are about 125,000 loci shared by the reference and study data.
    The golden standard error is the square root of the averate squared distance between the study individuals' PC scores predicted by different methods compared to the golden standard result, which in our case is the PC scores predicted by ADT.
  }
  \label{tbl:ukb}
\end{table}

\begin{table}
  \centering
  \begin{tabular}{|l|l|l|l|l|l|}
    \hline
    Reference Size & 1000 & 1500 & 2000 & 2500 & 3000 \\
    \hline
    ADT        &  68  & 1263 & 2871 & 5720 & 9737 \\
    OADT       &  72  & 18   & 19   & 22   & 24   \\
    ROADT      & 809  & 18   & 17   & 20   & 21   \\
    SP         &  68  & 26   & 42   & 67   & 96   \\
    RSP        & 805  & 23   & 31   & 39   & 46   \\
    AP         &  68  & 69   & 107  & 91   & 119  \\
    RAP        & 805  & 11   & 24   & 29   & 9    \\
    \hline
  \end{tabular}
  \caption{
    Reference runtimes for the simulated data as the reference size increases.
    The study size is fixed to 200 and the number of SNPs is 100,000 (1000 per genealogy). 
    Simulation is done on a $2 \times 2$ grid with a migration rate of 100 by the GGS software. 
  }
  \label{tbl:nChg-runtimes-ref}
\end{table}

\begin{table}
  \centering
  \begin{tabular}{|l|l|l|l|l|l|}
    \hline
    Reference Size & 1000 & 1500 & 2000 & 2500 & 3000 \\
    \hline
    ADT        & 391  &  130 &  222 &  339 &  486 \\
    OADT       & 15   &  136 &  230 &  350 &  499 \\
    ROADT      & 15   & 1211 & 1633 & 2023 & 2422 \\
    SP         & 16   &  130 &  222 &  339 &  486 \\
    RSP        & 15   & 1204 & 1624 & 2012 & 2409 \\
    AP         & 34   &  130 &  222 &  339 &  486 \\
    RAP        & 18   & 1204 & 1624 & 2012 & 2409 \\
    \hline
  \end{tabular}
  \caption{
    Study runtimes for simulated data as the reference size increases.
    The study size is fixed to 200 and the number of SNPs is 100,000 (1000 per genealogy). 
    Simulation is done on a $2 \times 2$ grid with a migration rate of 100 by the GGS software. 
  }
  \label{tbl:nChg-runtimes-study}
\end{table}

\begin{table}
  \centering
  \begin{tabular}{|l|l|l|l|l|l|}
    \hline
    Study Size & 1000  & 1500 & 2000 & 2500 & 3000 \\
    \hline
    ADT        & 0.00  & 0.00 & 0.00 & 0.00 & 0.00 \\
    OADT       & 0.23  & 0.15 & 0.11 & 0.08 & 0.07 \\
    ROADT      & 0.23  & 0.15 & 0.12 & 0.08 & 0.08 \\
    SP         & 6.00  & 4.34 & 3.43 & 2.75 & 2.34 \\
    RSP        & 6.00  & 4.34 & 3.43 & 2.75 & 2.34 \\
    AP         & 0.46  & 0.37 & 0.31 & 0.26 & 0.22 \\
    RAP        & 3.02  & 2.93 & 2.54 & 2.15 & 1.91 \\
    \hline
  \end{tabular}
  \caption{
    Golden standard error for the simulated data as the reference size increases.
    The study size is fixed to 200 and the number of SNPs is 100,000 (1000 per genealogy). 
    Simulation is done on a $2 \times 2$ grid with a migration rate of 100 by the GGS software. 
  }
  \label{tbl:nChg-accuracy-gold}
\end{table}

\begin{table}
  \centering
  \begin{tabular}{|l|l|l|l|l|l|}
    \hline
    Reference Size & 1000  & 1500 & 2000 & 2500 & 3000 \\
    \hline
    ADT        & 4.53  & 4.13 & 3.81 & 3.59 & 3.23 \\
    OADT       & 4.43  & 4.06 & 3.77 & 3.55 & 3.20 \\
    ROADT      & 4.44  & 4.06 & 3.76 & 3.55 & 3.20 \\
    SP         & 5.79  & 4.51 & 3.89 & 3.56 & 3.19 \\
    RSP        & 5.79  & 4.51 & 3.89 & 3.56 & 3.19 \\
    AP         & 4.59  & 4.10 & 3.76 & 3.57 & 3.24 \\
    RAP        & 4.23  & 3.87 & 3.55 & 3.38 & 3.07 \\
    \hline
  \end{tabular}
  \caption{
    Geographic center error for the simulated data as the reference size increases.
    The study size is fixed to 200 and the number of SNPs is 100,000 (1000 per genealogy). 
    Simulation is done on a $2 \times 2$ grid with a migration rate of 100 by the GGS software. 
  }
  \label{tbl:nChg-accuracy-ctr}
\end{table}

\begin{table}
  \centering
  \begin{tabular}{|l|l|l|l|l|l|}
    \hline
    Reference Size & 1000 & 1500 & 2000 & 2500 & 3000 \\
    \hline
    ADT        & 507  & 758 & 1011 & 1262 & 1549 \\
    OADT       &  73  & 108 &  156 &  189 &  282 \\
    ROADT      &  73  & 108 &  159 &  183 &  281 \\
    SP         &  44  &  64 &   84 &  103 &  131 \\
    RSP        &  43  &  63 &   83 &  101 &  129 \\
    AP         &  66  &  81 &  100 &  117 &  146 \\
    RAP        &  22  &  15 &   15 &   21 &   24 \\
    \hline
  \end{tabular}
  \caption{
    Study runtimes for the simulated data as the study size increases.
    The reference size is fixed to 600 and the number of SNPs is 100,000 (1000 per genealogy). 
    Simulation is done on a $2 \times 2$ grid with a migration rate of 100 by the GGS software. 
  }
  \label{tbl:mChg-runtimes-study}
\end{table}
